---
title: "Assignment 5 - Decision Trees"
author: "Charles Lang"
date: "November 9, 2016"
output: html_document
---
For this assignment we will be using data from the Assistments Intelligent Tutoring system. This system gives students hints based on how they perform on math problems. 

#Install & call libraries
```{r}
install.packages("party", "rpart")

library(rpart)
library(party)
```

## Part I
```{r}
D1 <- data.frame(intelligent_tutor)

```

##Classification Tree
First we will build a classification tree to predict which students ask a teacher for help, which start a new session, or which give up, based on whether or not the student completed a session (D1$complete) and whether or not they asked for hints (D1$hint.y). 
```{r}

c.tree <- rpart(action ~ hint.y + complete, method="class", data=D1) #Notice the standard R notion for a formula X ~ Y

#Look at the error of this tree
printcp(c.tree)

#Plot the tree
post(c.tree, file = "tree.ps", title = "Session Completion Action: 1 - Ask teacher, 2 - Start new session, 3 - Give up")

```
## Part II

#Regression Tree

We want to see if we can build a decision tree to help teachers decide which students to follow up with, based on students' performance in Assistments. We will create three groups ("teacher should intervene", "teacher should monitor student progress" and "no action") based on students' previous use of the system and how many hints they use. To do this we will be building a decision tree using the "party" package. The party package builds decision trees based on a set of statistical stopping rules.

#Take a look at our outcome variable "score"
```{r}
hist(D1$score)
```

#Create a categorical outcome variable based on student score to advise the teacher using an "ifelse" statement
```{r}
D1$advice <- ifelse(D1$score <=0.4, "intervene", ifelse(D1$score > 0.4 & D1$score <=0.8, "monitor", "no action"))
```

#Build a decision tree that predicts "advice" based on how many problems students have answered before, the percentage of those problems they got correct and how many hints they required
```{r}
score_ctree <- ctree(factor(advice) ~ prior_prob_count + prior_percent_correct + hints, D1)
```

#Plot tree
```{r}
plot(score_ctree)
```

Please interpret the tree, which two behaviors do you think the teacher should most closely pay attemtion to?

#The teacher should pay close attention to the students that request more than 12 hints and/or the students that have a prior % correct of less than or equal to 63%. In general, the greater the number of hints and the lower the prior % correct, the more intervention the student will need.

#Test Tree
Upload the data "intelligent_tutor_new.csv". This is a data set of a differnt sample of students doing the same problems in the same system. We can use the tree we built for the previous data set to try to predict the "advice" we should give the teacher about these new students. 

```{r}
#Upload new data

D2 <- data.frame(intelligent_tutor_new)

#Generate predicted advice for new students based on tree generated from old students

D2$prediction <- predict(score_ctree, D2)

``` 
## Part III
Compare the predicted advice with the actual advice that these students recieved. What is the difference between the observed and predicted results?

#The prediction model has students going through intervention and monitoring that actually don't need it. If they all scored 100%, then no action is needed. This could actually waste time students that don't need any action whatsover. On the other hand, looking at the data of prior % correct, # of prior questions answered and # of hints given, it doesn't seem likely that all of these students would then make 100% on the test. They obviously have different levels of ability. So perhaps the test was faulty? Seems there is more than meets the data here. 

```{r}
D2$advice <- ifelse(D2$score <=0.4, "intervene", ifelse(D2$score > 0.4 & D2$score <=0.8, "monitor", "no action"))


#We'd use this if we trusted the conclusions we made in part II: (but we don't :))
#D2$advice <- ifelse(D2$hints > 12, "intervene", ifelse(D2$hints < 8 & D2$hints > 12, "monitor", "no action"))


#D2$advice2 <- ifelse(D2$prior_percent_correct <= 0.6, "intervene", ifelse(D2$prior_percent_correct > 0.6 & D2$prior_percent_correct <=0.8, "monitor", "no action"))

score_ctree <- ctree(factor(prediction))

# this is not working well :( Gives solid boxes. No bueno. - Actaully it's bueno. it's fine, they all made 100%.

score_ctree <- ctree(factor(prediction) ~ prior_prob_count + prior_percent_correct + hints, D2)

plot(score_ctree)
```

